---
title: "8 - exercise"
format: 
  html: 
    embed-resources: true
editor: visual
---

Your job in this exercise is to run the same risk workflow on a ticker of your choice and report the numbers. Pick something that moves - a single stock rather than a broad ETF is fine, and it'll make the tails more interesting.

### 1. Load libraries

Load `tidyverse`, `tidyquant`, `PerformanceAnalytics`, and `scales`.

```{r}
#| label: load-libraries
#| warning: false
#| message: false


```

### 2. Grab 10 years of daily data for a single ticker

Pick any ticker. Compute daily arithmetic returns.

```{r}
#| label: grab-data
#| warning: false


```

### 3. Plot the return series

Make a simple line chart of daily returns. Label your axes in percent.

```{r}
#| label: plot-returns


```

### 4. Compute historical 95% VaR, expected shortfall, and parametric VaR

Produce a small tibble with all three. Assume a \$500,000 position so you can also quote dollar losses.

```{r}
#| label: var-es


```

### 5. Run a stress test

Find the worst 20-day rolling return in your history. Report the end date and the dollar loss on your \$500,000 position.

```{r}
#| label: stress-test


```

### 6. Run a Monte Carlo price-path simulation

10,000 paths, 252 days, normal assumption using your historical mu and sigma. Report:

-   mean terminal 1-year return
-   median terminal 1-year return
-   5th percentile terminal 1-year return
-   expected shortfall at 5%

```{r}
#| label: monte-carlo


```

### 7. One-paragraph interpretation

Below the code, in plain English, write two or three sentences telling a non-data-scientist what the risk profile of this stock looks like based on what you found. What would you tell them before they bought \$500K of it?

**ANSWER:**
